### 2025-08-19 actuarial.qmd
20250818 - additions

In general, many so-called experts are "pseudo-experts" who do not know the boundaries of their competence—that is, they do not know what they do not know; they have the illusion of validity of their predictions and are [overconfident](#sec-cognitiveBiasesOverconfidence) about their predictions [@Kahneman2011].
Yet, many people arrogantly proclaim to have predictive powers, including in low-validity environments such as fantasy football.
Indeed, pundits are more likely to be television guests if they are opinionated, clear, and (overly) confident and make big, bold predictions, because they are more entertaining and their predictions seem more compelling [even though they tend to be less accurate than individuals whose thinking is more complex and less decisive; @Kahneman2011; @Silver2012].
Consider sports pundits like Stephen A. Smith and Skip Bayless who make bold predictions with uber confidence.
[Optimism](#sec-cognitiveBiasesOptimism) and [(over)confidence](#sec-cognitiveBiasesOverconfidence) are valued by society [@Kahneman2011].
Nevertheless, true experts know their limits in terms of knowledge and ability to predict.

@@

- Computers can process lots of information simultaneously.
So can humans.
But computers can to an even greater degree.
- Computers are faster at making calculations.
- Given the same input, a formula will give the exact same result everytime.
Humans' judgment tends to be inconsistent both across raters and within rater across time, when trying to make judgments or predictions from complex information [@Kahneman2011].
As noted in @sec-reliabilityVsValidity, [reliability](#sec-reliability) sets the upper bound for [validity](#sec-validity), so unreliable judgments cannot be accurate (i.e., valid).
- Computations by computers are error-free (as long as the computations are programmed correctly).
- Computers' judgments will not be biased by fatigue or emotional responses.
- Computers' judgments will tend not to be biased in the way that humans' [cognitive biases](#sec-cognitiveBias) are.
Computers are less likely to be [overconfident](#sec-cognitiveBiasesOverconfidence) in their judgments.
- Computers can more accurately weight the set of predictors based on large data sets.
Humans tend to give too much weight to singular predictors.
Experts may attempt to be clever and to consider complex combinations of predictors, but doing so often reduces validity [@Kahneman2011].
Simple combinations of predictions often outperform more complex combinations [@Kahneman2011].

@@

Actuarial methods are particularly valuable for criterion-referenced assessment tasks, in which the aim is to predict specific events or outcomes [@Garb2019].
For instance, actuarial methods have shown promise in predicting violence, criminal recidivism, psychosis onset, course of mental disorders, treatment selection, treatment failure, suicide attempts, and suicide [@Garb2019].
Actuarial methods are especially important to use in low-validity environments (like fantasy football) in which there is considerable uncertainty and unpredictability [@Kahneman2011].

>>>>

In general, many so-called experts are "pseudo-experts" who do not know the boundaries of their competence—that is, they do not know what they do not know; they have the illusion of validity of their predictions and are [overconfident](#sec-cognitiveBiasesOverconfidence) about their predictions [@Kahneman2011].
A person's confidence is not a good indicator of their accuracy [@Kahneman2011].
Yet, many people arrogantly proclaim to have predictive powers, including in low-validity environments such as fantasy football.
Indeed, pundits are more likely to be television guests if they are opinionated, clear, and (overly) confident and make big, bold predictions, because they are more entertaining and their predictions seem more compelling [even though they tend to be less accurate than individuals whose thinking is more complex and less decisive; @Kahneman2011; @Silver2012].
Consider sports pundits like Stephen A. Smith and Skip Bayless who make bold predictions with uber confidence.
[Optimism](#sec-cognitiveBiasesOptimism) and [(over)confidence](#sec-cognitiveBiasesOverconfidence) are valued by society [@Kahneman2011].
Nevertheless, true experts know their limits in terms of knowledge and ability to predict.

@@

- Computers can process lots of information simultaneously.
So can humans.
But computers can to an even greater degree.
- Computers are faster at making calculations.
- Given the same input, a formula will give the exact same result everytime.
Humans' judgment tends to be inconsistent both across raters and within rater across time, when trying to make judgments or predictions from complex information [@Kahneman2011].
As noted in @sec-reliabilityVsValidity, [reliability](#sec-reliability) sets the upper bound for [validity](#sec-validity), so unreliable judgments cannot be accurate (i.e., valid).
- Computations by computers are error-free (as long as the computations are programmed correctly).
- Computers' judgments will not be biased by fatigue or emotional responses.
- Computers' judgments will tend not to be biased in the way that humans' [cognitive biases](#sec-cognitiveBias) are.
Computers are less likely to be [overconfident](#sec-cognitiveBiasesOverconfidence) in their judgments.
- Computers can more accurately weight the set of predictors based on large data sets.
Humans tend to give too much weight to singular predictors.
Experts may attempt to be clever and to consider complex combinations of predictors, but doing so often reduces validity [@Kahneman2011].
Simple combinations of predictions (e.g., with few predictors and/or predictors of equal weight) often outperform more complex combinations [@Dawes1979; @Kahneman2011].

@@

Actuarial methods are particularly valuable for criterion-referenced assessment tasks, in which the aim is to predict specific events or outcomes [@Garb2019].
For instance, actuarial methods have shown promise in predicting violence, criminal recidivism, psychosis onset, course of mental disorders, treatment selection, treatment failure, suicide attempts, and suicide [@Garb2019].
Actuarial methods are especially important to use in low-validity environments (like fantasy football) in which there is considerable uncertainty and unpredictability [@Kahneman2011].
Actuarial methods may outperform human judgment in low-validity environments because formulas are better able to detect weakly valid cues and to maintain a modest degree of accuracy using such cues consistently [@Kahneman2011].
